---
title: "Lo_Giudice_090321"
author: "Rachel Jackson"
date: "23/03/2021"
output: html_document
---

Load required libraries
```{r}
library(SingleCellExperiment)
library(reticulate)
use_python("/Applications/Python 3.7")
library(cellassign)
library(Matrix)
library(ggplot2)
library(scater)
library(scran)
library(DropletUtils)
library(PCAtools)
library(dplyr)

```

```{r}
sce = readRDS("LoGiudice_QualityControl_data_sce.rds")
sce
```
Split sce object into Lane 1 only:
```{r}
lane1_bool <- sce$lane == 'Lane1'
sce_lane1 <- sce[,lane1_bool]
sce_lane1
```

Generate Gene Ensembl IDs and Gene types:
```{r}
library(biomaRt)
library(dplyr)
genes<-rowData(sce_lane1)$genes
mart <- biomaRt::useMart(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")
t2g <- biomaRt::getBM(attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"), mart = mart)
t2g <- dplyr::rename(t2g, ens_gene = ensembl_gene_id, ext_gene = external_gene_name, biotype = gene_biotype)
# create vector of ensemble ids
gene_ens_ids <- c()
gene_type    <- c()
for (gene in genes) {
  if (gene %in% t2g$ext_gene){
    tmp_ens_id = t2g$ens_gene[t2g$ext_gene == gene]
    tmp_type   = t2g$biotype[t2g$ext_gene == gene]
    gene_ens_ids <- c(gene_ens_ids, tmp_ens_id[1]) # < a bit dodgey!
    gene_type    <- c(gene_type, tmp_type[1])      # < a bit dodgey!
  }
  else {
    gene_ens_ids <- c(gene_ens_ids, 'none')
    gene_type <- c(gene_type, 'none')
  }
}
length(gene_ens_ids)
length(gene_type)
length(genes)
```

Add Gene Ensembl IDs and Gene types to gene metadata:
```{r}
rowData(sce_lane1)$ensembl_gene_id<- gene_ens_ids
rowData(sce_lane1)$gene_type<- gene_type
sce_lane1 <- sce_lane1[rowData(sce_lane1)$gene_type=="protein_coding",]
```

Define markers for cell assign (from Clark et al.(2019)):
```{r}
marker_list <- list(
  `Photoreceptor precursors` = c("Opn1sw","Crx"),
  `Neurogenic` = c("Rax", "Dll1","Neurog2", "Hes6", "Olig2", "Dll4"),
  `Amacrine Cells` = c("Dlx1", "Sox11", "Pax6", "Sox4"),
  `Late Progenitor` = c("Hes5", "Id1"),
  `Intermediate Progenitor` = c("Heyl", "Sox8"),
  `Early Progenitor` = c("Fgf15", "Ccnd2", "Sfrp2"), 
  `RGC` =   c("Isl1", "Elavl4")
)
markers <- cellassign::marker_list_to_mat(marker_list)
markers
```

Compute and store sizefactors:
```{r}
set.seed(42)
quick_clusters <- quickCluster(sce_lane1)
sce_lane1<- computeSumFactors(sce_lane1, clusters = quick_clusters)
print(summary(sizeFactors(sce_lane1)))
sce_lane1<- logNormCounts(sce_lane1)
```

Apply `cellAssign` to label cells by markers
```{r}
sizefactors <- calculateSumFactors(sce_lane1)
fit_cells <- cellassign(exprs_obj = sce_lane1[rownames(markers),], 
                  marker_gene_info = markers, 
                  s = sizefactors, 
                  learning_rate = 1e-2, 
                  shrinkage = TRUE,
                  verbose = FALSE)

print(head(celltypes(fit_cells)))
print(head(cellprobs(fit_cells)))
```
t-SNE plot by CellAssign markers for Lane 1 only:
```{r, echo=TRUE}
sce_lane1$cell_type <- fit_cells$cell_type
set.seed(00101001101)
sce_lane1<- runTSNE(sce_lane1, exprs_values = "logcounts")
plotReducedDim(sce_lane1, dimred="TSNE", colour_by="cell_type") 
```

Run CytoTRACE (gives differentiation scores for each cell):
```{r, echo=TRUE}
library(CytoTRACE)
CytoTRACE_results <- CytoTRACE(as.matrix(counts(sce_lane1)), enableFast = TRUE, ncores = 2)
```
Add CytoTRACE scores to cell metadata:
```{r, echo=TRUE}
sce_lane1$CytoTRACE_results<- CytoTRACE_results$CytoTRACE
```

Plot of CellAssign and CytoTRACE scores:
```{r, echo=TRUE, fig1, fig.height = 5, fig.width = 12}
set.seed(00101001101)
a<-plotTSNE(sce_lane1, colour_by="cell_type")
b<-plotTSNE(sce_lane1, colour_by="CytoTRACE_results")
multiplot(a,b,cols=2)

```

Add Cell Cycle scores and phases to cell metadata:
```{r}
mouse.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))
assignments <- cyclone(sce_lane1, mouse.pairs, gene.names=rowData(sce_lane1)$ensembl_gene_id)
# record the G1, G2M and Phase scores for future reference
colData(sce_lane1)$cc_G1_score <- assignments$scores$G1
colData(sce_lane1)$cc_G2M_score <- assignments$scores$G2M
colData(sce_lane1)$cc_S_score <- assignments$scores$S
colData(sce_lane1)$cc_Phase <- assignments$phases
```

Transformations and HVG selection:
Create assays:
```{r}
assay(sce_lane1, "normcounts") <- exp(assay(sce_lane1, "logcounts")) - 1
assay(sce_lane1, "sqrt") <- sqrt(assay(sce_lane1, "normcounts"))
assay(sce_lane1, "FT") <- sqrt(assay(sce_lane1, "normcounts")) + sqrt(assay(sce_lane1, "normcounts") + 1)
```
Model the mean-variance relationship for each gene in the data, for each transformation of the data
```{r}
library(scry)
sce_lane1 <- devianceFeatureSelection(sce_lane1, assay="counts", sorted=TRUE)
MV_log <- modelGeneVar(sce_lane1, assay.type = "logcounts")
MV_sqrt <- modelGeneVar(sce_lane1, assay.type = "sqrt")
MV_FT <- modelGeneVar(sce_lane1, assay.type = "FT")
```
Get the top 2000/3000 Highly Variable Genes for each transformation and add to gene metadata:
```{r}
HVG_log <- getTopHVGs(MV_log, n = 2000, var.threshold = NULL)
HVG_sqrt <- getTopHVGs(MV_sqrt, n = 2000, var.threshold = NULL)
HVG_FT <- getTopHVGs(MV_FT, n = 2000, var.threshold = NULL)
HVG_dev <- rownames(sce_lane1)[1:2000]
HVG_log3000 <- getTopHVGs(MV_log, n = 3000, var.threshold = NULL)
HVG_sqrt3000 <- getTopHVGs(MV_sqrt, n = 3000, var.threshold = NULL)
HVG_FT3000 <- getTopHVGs(MV_FT, n = 3000, var.threshold = NULL)
HVG_dev3000 <- rownames(sce_lane1)[1:3000]
rowData(sce_lane1)$HVG_log <- ifelse(rownames(sce_lane1) %in% HVG_log, TRUE, FALSE)
rowData(sce_lane1)$HVG_sqrt <- ifelse(rownames(sce_lane1) %in% HVG_sqrt, TRUE, FALSE)
rowData(sce_lane1)$HVG_FT <- ifelse(rownames(sce_lane1) %in% HVG_FT, TRUE, FALSE)
rowData(sce_lane1)$HVG_dev <- ifelse(rownames(sce_lane1) %in% HVG_dev, TRUE, FALSE)
rowData(sce_lane1)$HVG_log3000 <- ifelse(rownames(sce_lane1) %in% HVG_log3000, TRUE, FALSE)
rowData(sce_lane1)$HVG_sqrt3000 <- ifelse(rownames(sce_lane1) %in% HVG_sqrt3000, TRUE, FALSE)
rowData(sce_lane1)$HVG_FT3000 <- ifelse(rownames(sce_lane1) %in% HVG_FT3000, TRUE, FALSE)
rowData(sce_lane1)$HVG_dev3000 <- ifelse(rownames(sce_lane1) %in% HVG_dev3000, TRUE, FALSE)
```

To save files:
```{r}
library(Matrix)
library(scater)
# save cell_metadata
write.csv(colData(sce_lane1),file="LoGiudice_cell_metadata_file_lane1_3")
# save gene metadata
write.csv(rowData(sce_lane1),file="LoGiudice_gene_metadata_file_lane1_3")
# save counts matrix
Matrix::writeMM(as(counts(sce_lane1), "dgTMatrix"), file='LoGiudice_counts_file_lane1_3' )
```




